---
title: "Putting the exponentials in the functions"
format: 
    html: 
        code-fold: true
---

## Preliminaries

```{r}
library(brms)
library(tidyverse)
```

Dr. Zoh's preliminary code. This time, though, we put the exponentials on the definition of the functions. 
```{r}
#-------------------------------------------------------------------------------
#--- Useful functions
#-- find the root
#-------------------------------------------------------------------------------
GamF <- function(gam, p0) {
          (2 * pnorm(-abs(gam)) * exp(.5 * gam^2) - p0)^2
}

# optimize(GamF,interval = c(-30,30), p0=1 - 0.05)
GamBnd <- function(p0) {
          Re1 <- optimize(GamF, interval = c(-30, 30), p0 = 1 - p0)
          Re2 <- optimize(GamF, interval = c(-30, 30), p0 = p0)

          c(-abs(Re1$minimum), abs(Re2$minimum), Re1$objective, Re2$objective)
}
GamBnd <- Vectorize(GamBnd, "p0")

#-------------------------------------------------------------------------------
#--- Function
#-------------------------------------------------------------------------------
#--- Stan function (case 2)
tau0 <- .1


Bd <- GamBnd(tau0)[1:2]
# GAL2 <- custom_family(
#  "GAL2", dpars = c("mu", "sigma","ligam", "tau","gamL","gamU"), links=c("identity","log","identity","identity", rep("identity",2)),
#  lb=c(NA,0, -1.087,0,NA,NA), ub=c(NA,NA, 1.087,1,NA,NA), type="real") #, vars = "vint1[n]"
# GAL2 <- custom_family(
#   "GAL2", dpars = c("mu", "sigma","ligam"), links= c("identity","log","identity"),
#   lb=c(NA,0, -1.087), ub=c(NA,NA, 1.087), type="real", vars = c("tau","gamL","gamU")) #, vars = "vint1[n]"
GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "sigma", "ligam", "tau"), links = c("identity", "log", "identity", "identity"),
          lb = c(NA, 0, -1.086, 0), ub = c(NA, NA, 1.086, 1), type = "real"
) # , vars = c("tau")) #, vars = "vint1[n]"

GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "sigma", "ligam", "tau"), links = c("identity", "log", "logit", "identity"),
          lb = c(NA, 0, -15, 0), ub = c(NA, NA, 15, 1), type = "real", vars = c("gamL", "gamU")
) # , vars = c("tau")) #, vars = "vint1[n]"

GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "sigma", "ligam", "tau", "gamL", "gamU"), links = c("identity", "log", "identity", "identity", rep("identity", 2)),
          lb = c(NA, 0, -15, 0, NA, NA), ub = c(NA, NA, 15, 1, NA, NA), type = "real"
) # , vars = "vint1[n]"

GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "sigma", "ligam", "tau"), links = c("identity", "log", "identity", "identity"),
          lb = c(NA, 0, Bd[1] * .99, 0), ub = c(NA, NA, Bd[2] * .99, 1), type = "real"
) # , vars = "vint1[n]"


stan_funs2 <- "
    /*
    A = -est*p_neg + .5*pow(gam, 2)*pow(p_neg/p_pos, 2) + log(Phi_approx(a2-a3)) + log1m_exp(fabs(log(Phi_approx(a2-a3)) - log(Phi_approx(a2))));
  gam = (gamU - gamL) * ligam + gamL;
  real gam = (gamU - gamL) * ligam + gamL;
  real GAL2_lpdf(real y, real mu, real sigma, real ligam, real tau, real gamL, real gamU){
  real GAL2_rng(real mu, real sigma, real ligam, real tau, real gamL, real gamU){
   */
     /* helper function for asym_laplace_lpdf
  * Args:
    *   y: the response value
  *   tau: quantile parameter in (0, 1)
  */
    real rho_quantile(real y, real tau) {
      if (y < 0) {
        return y * (tau - 1);
      } else {
        return y * tau;
      }
    }
  /* asymmetric laplace log-PDF for a single response
  * Args:
    *   y: the response value
  *   mu: location parameter
  *   sigma: positive scale parameter
  *   tau: quantile parameter in (0, 1)
  * Returns:
    *   a scalar to be added to the log posterior
  */
    real asym_laplace_lpdf(real y, real mu, real logsigma, real tau) {
      return log(tau * (1 - tau)) -
        logsigma -
        rho_quantile((y - mu) / exp(logsigma), tau);
    }
  /* asymmetric laplace log-CDF for a single quantile
  * Args:
    *   y: a quantile
  *   mu: location parameter
  *   sigma: positive scale parameter
  *   tau: quantile parameter in (0, 1)
  * Returns:
    *   a scalar to be added to the log posterior
  */
    real asym_laplace_lcdf(real y, real mu, real logsigma, real tau) {
      if (y < mu) {
        return log(tau) + (1 - tau) * (y - mu) / exp(logsigma);
      } else {
        return log1m((1 - tau) * exp(-tau * (y - mu) / exp(logsigma)));
      }
    }
  /* asymmetric laplace log-CCDF for a single quantile
  * Args:
    *   y: a quantile
  *   mu: location parameter
  *   sigma: positive scale parameter
  *   tau: quantile parameter in (0, 1)
  * Returns:
    *   a scalar to be added to the log posterior
  */
    real asym_laplace_lccdf(real y, real mu, real logsigma, real tau) {
      if (y < mu) {
        return log1m(tau * exp((1 - tau) * (y - mu) / exp(logsigma)));
      } else {
        return log1m(tau) - tau * (y - mu) / exp(logsigma);
      }
    }

   real GAL2_lpdf(real y, real mu, real logsigma, real ligam, real tau){

   real p_pos;
   real p_neg;
   real a3;
   real a2;
   real p;
   real est;
   real A;
   real B;
   real Res = 0;
   real gam = ligam;
    p = (gam < 0) + (tau - (gam < 0))/(2*Phi(-fabs(gam))*exp(.5*pow(gam, 2)));
    p_pos = p -  (gam > 0);
    p_neg = p -  (gam < 0);
    est = (y - mu) / exp(logsigma);

    if(fabs(gam) > 0){
    a3 = p_pos * (est / fabs(gam));
    a2 = fabs(gam) * (p_neg / p_pos);


    if(est/gam > 0){
      A =  0.5*pow(gam, 2)*pow(p_neg/p_pos, 2) - est*p_neg + log_diff_exp(log(Phi_approx(a2-a3)), log(Phi_approx(a2)) );
      B =  0.5*pow(gam, 2) - p_pos*est + log(Phi_approx(-fabs(gam) + a3));
      Res = log(2*p*(1-p)) - logsigma +  log_sum_exp(A, B);
    }else{
      Res =  log(2*p*(1-p)) - logsigma - p_pos * est + 0.5 * pow(gam, 2) + log(Phi_approx(-fabs(gam) ));
    }
    }else{
    Res = asym_laplace_lpdf( y | mu, exp(logsigma), tau);
    }

    return Res;
   }

  real GAL2_rng(real mu, real logsigma, real ligam, real tau){

     real A;
     real B;
     real C;
     real p;
     real hi;
     real nui;
     real mui=0;
     real Up = uniform_rng(.5, 1.0);

     real gam = ligam;
     p = (gam < 0) + (tau - (gam < 0))/(2*Phi_approx(-fabs(gam))*exp(.5*pow(gam, 2)));
     A = (1 - 2*p)/(p - pow(p,2));
     B = 2/(p - pow(p,2));
     C = 1/((gam > 0) - p);

      hi = exp(logsigma) * inv_Phi(Up);
     nui = exp(logsigma) * exponential_rng(1);
     mui += mu + A * nui + C * fabs(gam) * hi;

     return normal_rng(mui, sqrt(exp(logsigma)*B*nui));
  }
  "


#--- Now define all of these here
# stanvars <- stanvar(scode = stan_funs, block = "functions")
stanvars2 <- stanvar(scode = stan_funs2, block = "functions")

# priorval = c(prior(normal(0, 15), class = "Intercept"),
#           # Prior guess of 80% of the terms are non-zero
#           prior(horseshoe(par_ratio =0.8), class = "b"))

#--- Fit the quantile regression
#-- quantile

# fit_fakeq.5 <- brms::brm(bf(as.formula(form), quantile = .5), data = df_sampleWt, family = asym_laplace(), prior = set_prior(horseshoe(df=1,par_ratio = 0.1), class="b"))
# fit_fakeq.5 <- brms::brm(bf(as.formula(form), quantile = .5), data = df_sampleWt, family = asym_laplace())

# fit_fakeqGal.5 <- brms::brm(bf(as.formula(form), tau = .5, gamL = Bd[1], gamU = Bd[2]), data = df_sampleWt, family = GAL,
#                             stanvars = stanvars,chains = 3,iter = 10000, control = list(adapt_delta = 0.97, max_treedepth = 10))


# fit_fakeqGal2.5 <- brms::brm(bf(as.formula(form), tau = .5, gamL = Bd[1], gamU = Bd[2]), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 3,iter = 1000)
# fit_fakeqGal2.5 <- brms::brm(bf(as.formula(form), tau = .5,gamL = Bd[1], gamU = Bd[2]), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 2000, init = 0.1, control = list(adapt_delta = 0.97))

# fit_fakeqGal2.5 <- brms::brm(bf(as.formula(form), tau = tau0), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,
# fit_fakeqGal2.9 <- brms::brm(bf(as.formula(form), tau = tau0), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,
# fit_fakeqGal2.1 <- brms::brm(bf(as.formula(form), tau = tau0), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,

```


Generating synthethic data with mean 0 and $\mathrm{sd} = x$

```{r}
N <- 1000
x <- runif(N, 0, 1)
y <- rnorm(N, mean = 0, sd = x)

synthetic <- data.frame(x = x, y = y)
plot(x, y)
```

## Experimenting 

Original model 

```{r}
Bd <- GamBnd(.5)[1:2]
GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "logsigma", "ligam", "tau"), links = c("identity", "identity", "identity", "identity"),
          lb = c(NA, NA, Bd[1] * .99, 0), ub = c(NA, NA, Bd[2] * .99, 1), type = "real"
)
q50 <- brm(bf(y ~ x, logsigma ~ log(x) - 1, tau = .5), data = synthetic, family = GAL2, stanvars = stanvars2, chains = 2, iter = 500, control = list(adapt_delta = 0.99), cores = 4, seed = 123)
summary(q50)
```

```{r}
plot(q50)
```


```{r}
pairs(q50)
```



## Second attempt

Not fitting location

```{r}
Bd <- GamBnd(.5)[1:2]
GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "logsigma", "ligam", "tau"), links = c("identity", "identity", "identity", "identity"),
          lb = c(NA, NA, Bd[1] * .99, 0), ub = c(NA, NA, Bd[2] * .99, 1), type = "real"
)
q50 <- brm(bf(y ~ 1, logsigma ~ log(x) - 1, tau = .5), data = synthetic, family = GAL2, stanvars = stanvars2, chains = 2, iter = 500, control = list(adapt_delta = 0.99), cores = 4, seed = 123)
summary(q50)
```

```{r}
plot(q50)
```


```{r}
pairs(q50)
```


## Third attempt

not fitting location and not using logs on the RHS

```{r}
Bd <- GamBnd(.5)[1:2]
GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "logsigma", "ligam", "tau"), links = c("identity", "identity", "identity", "identity"),
          lb = c(NA, NA, Bd[1] * .99, 0), ub = c(NA, NA, Bd[2] * .99, 1), type = "real"
)
q50 <- brm(bf(y ~ 1, logsigma ~ x - 1, tau = .5), data = synthetic, family = GAL2, stanvars = stanvars2, chains = 2, iter = 500, control = list(adapt_delta = 0.99), cores = 4, seed = 123)
summary(q50)
```

```{r}
plot(q50)
```


```{r}
pairs(q50)
```


## Fourth attempt 

Not fitting location and fitting a smooth function

```{r}
Bd <- GamBnd(.5)[1:2]
GAL2 <- custom_family(
          "GAL2",
          dpars = c("mu", "logsigma", "ligam", "tau"), links = c("identity", "identity", "identity", "identity"),
          lb = c(NA, NA, Bd[1] * .99, 0), ub = c(NA, NA, Bd[2] * .99, 1), type = "real"
)
q50 <- brm(bf(y ~ 1, logsigma ~ s(x), tau = .5), data = synthetic, family = GAL2, stanvars = stanvars2, chains = 2, iter = 500, control = list(adapt_delta = 0.99), cores = 4, seed = 123)
summary(q50)
```

```{r}
plot(q50)
```


```{r}
pairs(q50)
```


```{r}
msms <- conditional_smooths(q50)
plot(msms)
```


