---
title: "Posterior Predictive Checks"
format: 
    html: 
        code-fold: true
---

```{r}
library(brms)
library(tidyverse)
```

Dr. Zoh's preliminary code:

```{r}
#-------------------------------------------------------------------------------
#--- Useful functions
#-- find the root
#-------------------------------------------------------------------------------
GamF <- function(gam, p0){
  
  (2*pnorm(-abs(gam))*exp(.5*gam^2) - p0)^2
}

#optimize(GamF,interval = c(-30,30), p0=1 - 0.05)
GamBnd <- function(p0){
  
  Re1 <- optimize(GamF, interval = c(-30, 30), p0 = 1-p0)
  Re2 <- optimize(GamF, interval = c(-30, 30), p0 = p0)
  
  c( -abs(Re1$minimum),  abs(Re2$minimum), Re1$objective,Re2$objective)  
}
GamBnd <- Vectorize(GamBnd,"p0")

#-------------------------------------------------------------------------------
#--- Function
#-------------------------------------------------------------------------------
#--- Stan function (case 2)
tau0 = .1

{
  Bd = GamBnd(tau0)[1:2]
  #GAL2 <- custom_family(
  #  "GAL2", dpars = c("mu", "sigma","ligam", "tau","gamL","gamU"), links=c("identity","log","identity","identity", rep("identity",2)), 
  #  lb=c(NA,0, -1.087,0,NA,NA), ub=c(NA,NA, 1.087,1,NA,NA), type="real") #, vars = "vint1[n]"
  # GAL2 <- custom_family(
  #   "GAL2", dpars = c("mu", "sigma","ligam"), links= c("identity","log","identity"), 
  #   lb=c(NA,0, -1.087), ub=c(NA,NA, 1.087), type="real", vars = c("tau","gamL","gamU")) #, vars = "vint1[n]"
  GAL2 <- custom_family(
    "GAL2", dpars = c("mu", "sigma","ligam", "tau"), links= c("identity","log","identity","identity"), 
    lb=c(NA,0, -1.086,0), ub=c(NA,NA, 1.086,1), type="real") #, vars = c("tau")) #, vars = "vint1[n]"
  
  GAL2 <- custom_family(
    "GAL2", dpars = c("mu", "sigma","ligam", "tau"), links= c("identity","log","logit","identity"), 
    lb=c(NA,0, -15,0), ub=c(NA,NA, 15,1), type="real", vars=c("gamL", "gamU")) #, vars = c("tau")) #, vars = "vint1[n]"
  
  GAL2 <- custom_family(
    "GAL2", dpars = c("mu", "sigma","ligam", "tau","gamL","gamU"), links=c("identity","log","identity","identity", rep("identity",2)),
    lb=c(NA,0, -15,0,NA,NA), ub=c(NA,NA, 15,1,NA,NA), type="real") #, vars = "vint1[n]"
  
  GAL2 <- custom_family(
    "GAL2", dpars = c("mu", "sigma","ligam", "tau"), links=c("identity","log","identity","identity"),
    lb=c(NA,0, Bd[1]*.99,0), ub=c(NA,NA, Bd[2]*.99,1), type="real") #, vars = "vint1[n]"
  
  
  stan_funs2 <- "
  /*
  A = -est*p_neg + .5*pow(gam, 2)*pow(p_neg/p_pos, 2) + log(Phi_approx(a2-a3)) + log1m_exp(fabs(log(Phi_approx(a2-a3)) - log(Phi_approx(a2)))); 
  gam = (gamU - gamL) * ligam + gamL;
  real gam = (gamU - gamL) * ligam + gamL;
  real GAL2_lpdf(real y, real mu, real sigma, real ligam, real tau, real gamL, real gamU){
  real GAL2_rng(real mu, real sigma, real ligam, real tau, real gamL, real gamU){
   */
     /* helper function for asym_laplace_lpdf
  * Args:
    *   y: the response value
  *   tau: quantile parameter in (0, 1)
  */
    real rho_quantile(real y, real tau) {
      if (y < 0) {
        return y * (tau - 1);
      } else {
        return y * tau;
      }
    }
  /* asymmetric laplace log-PDF for a single response
  * Args:
    *   y: the response value
  *   mu: location parameter
  *   sigma: positive scale parameter
  *   tau: quantile parameter in (0, 1)
  * Returns:
    *   a scalar to be added to the log posterior
  */
    real asym_laplace_lpdf(real y, real mu, real sigma, real tau) {
      return log(tau * (1 - tau)) -
        log(sigma) -
        rho_quantile((y - mu) / sigma, tau);
    }
  /* asymmetric laplace log-CDF for a single quantile
  * Args:
    *   y: a quantile
  *   mu: location parameter
  *   sigma: positive scale parameter
  *   tau: quantile parameter in (0, 1)
  * Returns:
    *   a scalar to be added to the log posterior
  */
    real asym_laplace_lcdf(real y, real mu, real sigma, real tau) {
      if (y < mu) {
        return log(tau) + (1 - tau) * (y - mu) / sigma;
      } else {
        return log1m((1 - tau) * exp(-tau * (y - mu) / sigma));
      }
    }
  /* asymmetric laplace log-CCDF for a single quantile
  * Args:
    *   y: a quantile
  *   mu: location parameter
  *   sigma: positive scale parameter
  *   tau: quantile parameter in (0, 1)
  * Returns:
    *   a scalar to be added to the log posterior
  */
    real asym_laplace_lccdf(real y, real mu, real sigma, real tau) {
      if (y < mu) {
        return log1m(tau * exp((1 - tau) * (y - mu) / sigma));
      } else {
        return log1m(tau) - tau * (y - mu) / sigma;
      }
    }
   
   real GAL2_lpdf(real y, real mu, real sigma, real ligam, real tau){
   
   real p_pos;
   real p_neg;
   real a3;
   real a2;
   real p;
   real est;
   real A;
   real B;
   real Res = 0;
   real gam = ligam;
    p = (gam < 0) + (tau - (gam < 0))/(2*Phi(-fabs(gam))*exp(.5*pow(gam, 2)));
    p_pos = p -  (gam > 0);
    p_neg = p -  (gam < 0);  
    est = (y - mu) / sigma; 
    
    if(fabs(gam) > 0){
    a3 = p_pos * (est / fabs(gam));
    a2 = fabs(gam) * (p_neg / p_pos);
    
    
    if(est/gam > 0){
      A =  0.5*pow(gam, 2)*pow(p_neg/p_pos, 2) - est*p_neg + log_diff_exp(log(Phi_approx(a2-a3)), log(Phi_approx(a2)) ); 
      B =  0.5*pow(gam, 2) - p_pos*est + log(Phi_approx(-fabs(gam) + a3));
      Res = log(2*p*(1-p)) - log(sigma) +  log_sum_exp(A, B);
    }else{
      Res =  log(2*p*(1-p)) - log(sigma) - p_pos * est + 0.5 * pow(gam, 2) + log(Phi_approx(-fabs(gam) ));
    }
    }else{
    Res = asym_laplace_lpdf( y | mu, sigma, tau); 
    }
     
    return Res;
   }
  
  real GAL2_rng(real mu, real sigma, real ligam, real tau){
  
     real A;
     real B;
     real C;
     real p;
     real hi;
     real nui;
     real mui=0;
     real Up = uniform_rng(.5, 1.0);
     
     real gam = ligam;
     p = (gam < 0) + (tau - (gam < 0))/(2*Phi_approx(-fabs(gam))*exp(.5*pow(gam, 2)));
     A = (1 - 2*p)/(p - pow(p,2));
     B = 2/(p - pow(p,2));
     C = 1/((gam > 0) - p);
     
      hi = sigma * inv_Phi(Up);
     nui = sigma * exponential_rng(1);
     mui += mu + A * nui + C * fabs(gam) * hi;
  
     return normal_rng(mui, sqrt(sigma*B*nui));
  }
  "
}  

#--- Now define all of these here
# stanvars <- stanvar(scode = stan_funs, block = "functions")
stanvars2 <- stanvar(scode = stan_funs2, block = "functions")

# priorval = c(prior(normal(0, 15), class = "Intercept"),
#           # Prior guess of 80% of the terms are non-zero
#           prior(horseshoe(par_ratio =0.8), class = "b"))

#--- Fit the quantile regression
#-- quantile

#fit_fakeq.5 <- brms::brm(bf(as.formula(form), quantile = .5), data = df_sampleWt, family = asym_laplace(), prior = set_prior(horseshoe(df=1,par_ratio = 0.1), class="b"))
#fit_fakeq.5 <- brms::brm(bf(as.formula(form), quantile = .5), data = df_sampleWt, family = asym_laplace())

# fit_fakeqGal.5 <- brms::brm(bf(as.formula(form), tau = .5, gamL = Bd[1], gamU = Bd[2]), data = df_sampleWt, family = GAL, 
#                             stanvars = stanvars,chains = 3,iter = 10000, control = list(adapt_delta = 0.97, max_treedepth = 10))


#fit_fakeqGal2.5 <- brms::brm(bf(as.formula(form), tau = .5, gamL = Bd[1], gamU = Bd[2]), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 3,iter = 1000)
#fit_fakeqGal2.5 <- brms::brm(bf(as.formula(form), tau = .5,gamL = Bd[1], gamU = Bd[2]), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 2000, init = 0.1, control = list(adapt_delta = 0.97))

#fit_fakeqGal2.5 <- brms::brm(bf(as.formula(form), tau = tau0), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,
#fit_fakeqGal2.9 <- brms::brm(bf(as.formula(form), tau = tau0), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,
# fit_fakeqGal2.1 <- brms::brm(bf(as.formula(form), tau = tau0), data = df_sampleWt, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,

```


Defining function required for posterior predictive 




## Case 1: homoskedastic, normally distributed data

Data generation:

```{r}
N <- 100
# mean number of data points per value of x sampled
mean_obs <- 30
# true value of beta, as in mu_i = beta * x_i
beta_mean <- 1
# multiplier for heteroskedasticity
hetero_fact <- 1


# simulation procedure
x <- c()
y <- c()
for (i in 1:N) {
    # generate x value
    x_i <- runif(n = 1)
    # generate how many samples will have that x value
    n_i <- rpois(n = 1, lambda = mean_obs)
    if (n_i > 0) {
        # generate the y values of each data point
        y_i <- rnorm(
            n = n_i, mean = beta_mean * x_i,
            sd = .1
            # sd = .1
        )
        # append them to data
        for (j in 1:n_i) {
            x <- append(x, x_i)
            y <- append(y, y_i[j])
        }
    }
}

simulated_data <- data.frame(x = x, y = y)
```


Fitting the model

```{r}
#| echo: false
#| message: false
#| warning: false
fit_fakeqGal2.1 <- brms::brm(bf(as.formula("y ~ x"), tau = tau0), data = simulated_data, family = GAL2, stanvars = stanvars2,chains = 2,iter = 5000, control = list(adapt_delta = 0.97)) # init = 0.1,
```


```{r}

summary(fit_fakeqGal2.1)


#brms::conditional_smooths(fit_fakeqGal2.5)
#brms::conditional_smooths(fit_fakeqGal2.9)
# brms::conditional_smooths(fit_fakeqGal2.1)

```

Additional function definition necessary for posterior predictive 

```{r}
posterior_predict_GAL2 <- function(i, prep, ...){
  mu <- brms::get_dpar(prep, 'mu', i = i)
  sigma <- brms::get_dpar(prep, 'sigma', i = i)
  ligam <- brms::get_dpar(prep, 'ligam', i = i)
  tau <- brms::get_dpar(prep, 'tau', i = i)
  GAL2_rng(mu, sigma, ligam, tau)
}
```

Making functions defined in preliminary code available for posterior predictive 

```{r}
#| echo: false
#| message: false
#| warning: false
expose_functions(fit_fakeqGal2.1, vectorize = TRUE)
```

generating draws from the posterior predictive for arbitrary values of $x$

```{r}
new_data <- data.frame(x = x[c(100, 500)])
post_draws <- posterior_predict(fit_fakeqGal2.1, new_data) 
```

plotting to perform check 

```{r}
comparison_data_1 <- simulated_data %>% filter(x == x[100])
comparison_data_2 <- simulated_data %>% filter(x == x[500])

par(mfrow = c(1,2))

plot(density(comparison_data_1[,2]), main = 'x = x[100]', ylim = c(0, 6))
lines(density(post_draws[,1]), col = 'blue')
legend('topright', c('actual', 'post pred'), col = c('black', 'blue'), pch = c(20))

plot(density(comparison_data_2[,2]), main = 'x = x[500]',  ylim = c(0,6))
lines(density(post_draws[,2]), col = 'blue')
legend('topright', c('actual', 'post pred'), col = c('black', 'blue'), pch = c(20))


```


